Summary Statistics

### 2025-03-13 extensive_tests/draft-sumstat-vignette.Rmd
a few little changes regarding chi-square test null hypothesis

A cherry $c$ is defined as a pair of direct offspring (tips \( t_1 \) and \( t_2 \)) 
of the same internal node in the phylogenetic tree. 
For a given cherry $c$, let $\mathcal{I}$ be the set of indices for island structures. 
For each island structure $i \in \mathcal{I}$ at each cherry tip $t_c \in {t_{c,1}, t_{c,2}$ in each phylogenetic tree cherries, 
the number of sites in each 
methylation state (unmethylated \( 0 \), partially-methylated \( 0.5 \), 
methylated \( 1 \)) is counted. This count is represented as:

@@

The distribution of methylation states at each island between the two tips \( t_1 \) 
and \( t_2 \) of a given cherry is compared using a chi-squared test. 
The null hypothesis \( H_0 \) is that the distributions of methylation states at 
the two tips are the same. The test statistic is calculated using the contingency 
table \( T \) of the methylation state counts for each tip:

$$
H_0: \text{distribution}(t_{c,1}) = \text{distribution}(t_{c,2})
$$

The contingency table \( T_{c,i} \) for tips \( t_{c,1} \) and \( t_{c,2} \) is:

@@

The p-value for the chi-squared test is calculated via Monte Carlo simulation to 
improve reliability, even when the expected frequencies do not meet the assumptions 
of the chi-squared test (i.e., expected counts of at least 5 in each category).

For each cherry $c$ and each island $i$, the methylation data at each cherry tip 
for the specified genomic structures (islands) are compared using the chi-squared test.

$$
p_{\text{value}}(c,i) = \chi^2\text{-test}(T_{c,i})
$$

These p-values are stored in a dataframe for each cherry and for each island.

>>>>

A cherry $c$ is defined as a pair of direct offspring (tips \( t_1 \) and \( t_2 \)) 
of the same internal node in the phylogenetic tree. 
For a given cherry $c$, let $\mathcal{I}$ be the set of indices for island structures. 
For each island structure $i \in \mathcal{I}$ at each cherry tip
$t_c \in \{t_{c,1}, t_{c,2}\}$ in each phylogenetic tree, 
the number of sites in each 
methylation state (unmethylated \( 0 \), partially-methylated \( 0.5 \), 
methylated \( 1 \)) is counted. This count is represented as:

@@

The distribution of methylation states at each island between the two tips \( t_1 \) 
and \( t_2 \) of a given cherry is compared using a chi-squared test. 
The null hypothesis of the test is that the frequency distributions of states
in the two islands follow the same multinomial distribution.
Note that not only IWEs can lead to deviations from this null hypothesis but
also neighbor-dependent SSEs can lead to slight deviations from the assumption
of multinomial distributions.
The test statistic is calculated using the contingency 
table \( T_{c,i} \) for tips \( t_{c,1} \) and \( t_{c,2} \):

@@

The p-value for the chi-squared test is calculated via Monte Carlo simulation to 
improve reliability, even when the expected frequencies do not meet the assumptions 
of the chi-squared approximation (i.e., expected counts of at least 5 in each category).

For each cherry $c$ and each island $i$, the methylation data at each cherry tip 
for the specified genomic structures (islands) are compared using the chi-squared test.
The obtained p-values are stored in a dataframe for each cherry and for each island.
